Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-01-15 2024pipit/section/70-Related.tex
related

Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
Compiling the triggers array with an unverified compiler such as Lustre V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
The Kind2 model-checker does include experimental array and quantifier support~\cite{kind2userdoc}, but uses a custom syntax for arrays with no compiler support.
This experimental support has some limitations: the use of arrays and quantifiers disables IC3-based invariant generation, quantified variables cannot be used in function calls, and top-level arrays are not supported~\cite{kind2024toparray}.

Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous} and JKind~\cite{gacek2018jkind} do not support quantifiers.
It may be possible to encode the quantifiers as fixed-size loops, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.

\TODO{Verified compilation via Frama-C} \cite{brun2023equation}

In terms of model-checking reactive systems,
% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.

The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}.
Recent work has used translation validation to show that the generated C code matches the high-level semantics~\cite{scott2023trustworthy}.
However, the model-checking suffers from the same semantic gap.

Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.

@@

\subsection{Model-checking, counterexamples and invariant generation}
\TODO{integration with model-checker}
Model-checkers do have distinct advantages over interactive theorem-provers: they are more usable, can find strengthening invariants automatically, and provide counterexamples.
However, they are limited to comparatively simple properties.
The time-triggered example from \autoref{s:motivation} would be difficult to encode in a model-checker with a limited specification language, as the trigger invariant uses an auxiliary recursive function to find the next active trigger.

In future work, we would like to explore integrating Pipit with an existing model-checker.
For simple first-order nonrecursive programs and specifications that fit within the bounds of another model-checker, we could generate an encoding to use for counterexample generation.
We would also like to investigate automatically lifting the strengthened invariants and suggesting the corresponding Pipit invariant.
Such a translation and lifting would not form part of the trusted computing base, as generated invariants would still be proved separately in \fstar{}.
In this way, we hope that it may be possible to obtain some of the usability benefits of model-checkers, with the added power of manual proofs when necessary, while retaining trust that the compiled code satisfies the original specification.

% \subsection{Model-checking, counterexamples and invariant generation}
\subsection{Comparison to a combinator-based verification approach}
\TODO{compare to refinement types, verification conditions, etc here}

>>>>

% SMT-based model-checkers~\cite{hagen2008scaling}

Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.
Recent work on translation validation for LustreC~\cite{brun2023equation} also does not yet support arrays.

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but is limited to verifying small arrays up to 8 or 16 triggers.
Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in multiple copies of the entire triggers array on the stack.

Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous}, JKind~\cite{gacek2018jkind} and the original Kind~\cite{hagen2008scaling} do not support quantifiers.
It may be possible to encode the quantifiers as fixed-size loops, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.

These model-checkers have definite usability advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
However, even when the problem can be expressed, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
In the future, we would like to investigate integrating Pipit with a model-checker via an unverified extraction: such an extraction may allow some of the usability benefits such as counterexamples and invariant generation.
If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not expand the trusted computing base.

Recent work has also introduced a form of refinement types for Lustre~\cite{chen2022synchronous}.
Rather than using transition systems, this work generates self-contained verification conditions based on the types of streams.
Such a type-based approach promises to allow abstraction of the implementation details.
However, for general-purpose functions such as \emph{count_when} from \autoref{s:motivation}, it is not clear how to give it a specification that actually \emph{abstracts} the implementation: a simple specification that the result is within some range would hide be insufficient for verifying the rest of the system.
For this function, the best specification is likely to include a re-statement of the implementation itself.

The embedded language Copilot generates real-time C code for runtime monitoring~\cite{laurent2015assuring}.
Recent work has used translation validation to show that the generated C code matches the high-level semantics~\cite{scott2023trustworthy}.
Copilot supports model-checking via Kind2; however, the model-checking has a limited specification language and does not support contracts.

Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs.
Compared to Vélus alone, the trusted computing base of Pipit is larger: we depend on all of \fstar{}, \lowstar{}'s C code extraction and the Z3 SMT solver.

@@

% \subsection{Comparison to a combinator-based verification approach}
% \TODO{compare to refinement types, verification conditions, etc here}
